Tree establishment in models of forest dynamics: a key priority for further research

### 2023-04-21 manuscript/manuscriptRW.Rmd
update acknowlegments with CH information
changes in the manuscript based on the comments from Mats and Christopher

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred J. Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Despite sometimes considerable mismatches between simulation results and
data, it is remarkable that most of the models capture the essential
features of the highly complex process of tree establishment, while not
having been parameterized with such data. We conclude that in spite of
all the uncertainties, a lot can be gained by a focus on the modeling of
establishment processes in these models. This has the potential to
render long-term projections of forest dynamics under changing
environmental conditions much more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change and increasing disturbances, it is becoming ever more
important to understand how forest structure, composition and function
will change, and to evaluate forest capacity to adapt to or be resilient
after disturbance [@lindner2010; @seidl2022]. A wide range of models of
forest dynamics were developed over the past decades considering the
impacts of climate on forest dynamics [@bugmann2022]. From these
studies, it is evident that we have a reasonably good understanding how
to model tree growth [@vanclay1997; @bugmann1996], and substantial
efforts have been dedicated to improving the representation of tree
mortality [@bugmann2019; @cailleret2017]. In contrast, tree
establishment is a much less studied process that is often represented
rather coarsely in these models [@walck2011; @leishman1992; @price2001],
which is a clear research gap in the context of climate-induced forest
disturbances and forest resilience.

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, climatic
water balance, soil quality and exposition (i.e., slope and aspect).
Each of these sites featured at least two consecutive measurements.
Ingrowth thresholds for the sites differed between diameters of 0 and 10
cm. For our study, we defined two data sets where one included 165 sites
with a diameter threshold of 7 cm or lower, and the other included
another 35 sites with diameter thresholds between 7 and 10 cm.

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol [@díaz-yáñez]
with instructions how to perform the simulations, providing input
variables on climate and soil conditions, and the expected output
variables. Neither were further site information (except for the data
specified below) nor any data on tree ingrowth or forest stand features
provided prior to the simulation. That is, the models were run in "blind
flight" mode.

@@

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 90 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management, due to misrepresentations in
density-dependent mortality processes in long-term simulations and
assumptions of tree geometry that lead to unrealistic single tree
dimensions of very old individuals. LPJ-GUESS had 2% of the sample
outputs without tree establishment, and these were considered as zero
stand basal area and zero ingrowth for all the tree species simulated;
one site produced grassland rather than a forest and was not included in
the results.

@@

```{=tex}
\begin{equation}
H_{R_{n}} = - \sum_{i = 1}^{s}{p_{R_{n,i}}\ln(p_{R_{n,i}})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{R_{n,i}} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{S_{n}} = - \sum_{i = 1}^{s}{p_{S_{n,i}}\ln(p_{S_{n,i}})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{S_{n,i}} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and that have a basal
area larger than zero in sample $n$; $p_{n,r,i}$ is the proportion of
the species $i$ in sample $n$ calculated as the ingrowth basal area
($r_{BA_{n,i}}$) for that species $i$ relative to the total recruited
basal area ($TotalR_{BA_{n}}$) of the sample $n$; $p_{S_{n,i}}$ is the
proportion of the species $i$ calculated as the basal area of all trees
($BA_{n,i}$) of that species $i$ relative to the total basal area
($TotalR_{BA_{n}}$) of the sample $n$.

@@

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
We assessed model preformance in relation to model traits focusing on
(1) model complexity as defined by @bugmann2022, (2) model type
(empirical or process based), (3) the presence of absence of a canopy
feedback for ingrowth, and (4) the scale of application of the model
(stand, landscape or global). We tested for significant differences
using two-sided t-tests by considering each of these model features
regarding the observed and simulated ingrowth levels and species
diversity, respectively.

@@

Lastly, there is no model that performed well across the five species
regarding the species-specific ingrowth levels in the environmental
space of Figure \@ref(fig:diff)). Some models (e.g., ForClim11, PICUS)
tended to consistently overestimate the ingrowth share of all five
species, but most models overestimated the ingrowth share of some
species while underestimating the share of others (e.g., ForCEEPS,
LPJ-GUESS).

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021 found
higher ingrowth rates than the ones observed in the Euforia data. Thus,
the overestimate of ingrowth rates by the models may not be so
problematic. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This may at
least partly explain e.g., the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

This study complies with the basic rules of reproducibility and has made
available the protocol of the study, the data and all the software that
has led to the findings presented in this paper. The protocol is
available here @díaz-yáñez . The code and data that support the findings
of this study are openly available at zenodo (*link to be added*) and
gihub (*link to be added*).

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project 200020_188882 / 1 . KB received support from
the European Research Council under the European Union's Horizon 2020
research and innovation program (Grant Agreement 101001905). JB
acknowledges funding from the long-term research development project no.
RVO 67985939 from the Czech Academy of Sciences. MM acknowledges
financial support from I-Maestro (Innovative forest management
strategies for a resilient bioeconomy under climate change and
disturbances, grant no. 22035418, 2019--2022) funded by the ERA-NET
Cofund ForestValue. KM received funding from the Project No.
CZ.02.1.01/0.0/0.0/16_019/0000803 financed by OP RDE, grant no.
ITMS2014+313011W580 supported by the Integrated Infrastructure
Operational Programme funded by the ERDF.

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred J. Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Christopher P.O. Reyer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Despite considerable mismatches between simulation results and data in
some instances, it is remarkable that most of the models capture the
essential features of the highly complex process of tree establishment,
while not having been parameterized with such data. We conclude that in
spite of all the uncertainties, a lot can be gained by a focus on the
modeling of establishment processes in these models. This has the
potential to render long-term projections of forest dynamics under
changing environmental conditions much more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change, increasing disturbances and changing societal demands on
forests ecosystems services, it is becoming ever more important to
understand how forest structure, composition and function will change,
and to evaluate forest capacity to adapt to or be resilient after
disturbance [@lindner2010; @seidl2022]. A wide range of models of forest
dynamics were developed over the past decades considering the impacts of
climate on forest dynamics [@bugmann2022]. From these studies, it is
evident that we have a reasonably good understanding how to model tree
growth [@vanclay1997; @bugmann1996], and substantial efforts have been
dedicated to improving the representation of tree mortality
[@bugmann2019; @cailleret2017]. In contrast, tree establishment is a
much less studied process that is often represented rather coarsely in
these models [@walck2011; @leishman1992; @price2001], which is a clear
research gap in the context of climate-induced forest disturbances and
forest resilience.

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, climatic
water balance, soil quality and exposition (i.e., slope and aspect).
Each of these sites featured at least two consecutive measurements.
Ingrowth thresholds for the sites differed between diameters of 0 and 10
cm. For our study, we defined two datasets where one included 165 sites
with a diameter threshold of 7 cm or lower, and the other included
another 35 sites with diameter thresholds between 7 and 10 cm.

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol
[@díaz-yáñez2022] with instructions how to perform the simulations,
providing input variables on climate and soil conditions, and the
expected output variables. Neither were further site information (except
for the data specified below) nor any data on tree ingrowth or forest
stand features provided prior to the simulation. That is, the models
were run in "blind flight" mode.

@@

Two models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 90 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management, due to misrepresentations in
density-dependent mortality processes in long-term simulations and
assumptions of tree geometry that lead to unrealistic single tree
dimensions of very old individuals. LPJ-GUESS had 2% of the sample
outputs without tree establishment, and these were considered as zero
stand basal area and zero ingrowth for all the tree species simulated;
one site produced grassland rather than a forest and was not included in
the results.

@@

```{=tex}
\begin{equation}
H_{R_{n}} = - \sum_{i = 1}^{s}{p_{R_{n,i}}\ln(p_{R_{n,i}})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{R_{n,i}} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{S_{n}} = - \sum_{i = 1}^{s}{p_{S_{n,i}}\ln(p_{S_{n,i}})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{S_{n,i}} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and that have a basal
area larger than zero in sample $n$; $p_{n,r,i}$ is the proportion of
the species $i$ in sample $n$ calculated as the ingrowth basal area
($r_{BA_{n,i}}$) for that species $i$ relative to the total recruited
basal area ($TotalR_{BA_{n}}$) of the sample $n$; $p_{S_{n,i}}$ is the
proportion of the species $i$ calculated as the basal area of all trees
($BA_{n,i}$) of that species $i$ relative to the total basal area
($TotalR_{BA_{n}}$) of the sample $n$. The species diversity was not
assessed for the model aDGVM2 because it does not simulate individual
species.

@@

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
We assessed model performance in relation to model traits focusing on
(1) model complexity as defined by @bugmann2022, (2) model type
(empirical or process based), (3) the presence of absence of a canopy
feedback for ingrowth, and (4) the scale of application of the model
(stand, landscape or global). We tested for significant differences
using two-sided t-tests by considering each of these model features
regarding the observed and simulated ingrowth levels and species
diversity, respectively.

@@

Lastly, there is no model that performed well across the five species
regarding the species-specific ingrowth levels in the environmental
space (Figure \@ref(fig:diff)). Some models (e.g., ForClim11, PICUS)
tended to consistently overestimate the ingrowth share of all five
species, but most models overestimated the ingrowth share of some
species while underestimating the share of others (e.g., ForCEEPS,
LPJ-GUESS).

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021 found
higher ingrowth rates than the ones observed in the EuFoRia data. Thus,
the overestimation of ingrowth rates by the models may not be so
problematic. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This may at
least partly explain e.g., the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

This study complies with the basic rules of reproducibility and has made
available the protocol of the study, the data and all the software that
has led to the findings presented in this paper. The protocol is
available here @díaz-yáñez2022 . The code and data that support the
findings of this study are openly available at zenodo (*link to be
added*) and gihub (*link to be added*).

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project 200020_188882 / 1 . KB received support from
the European Research Council under the European Union's Horizon 2020
research and innovation program (Grant Agreement 101001905). JB
acknowledges funding from the long-term research development project no.
RVO 67985939 from the Czech Academy of Sciences. MM acknowledges
financial support from I-Maestro (Innovative forest management
strategies for a resilient bioeconomy under climate change and
disturbances, grant no. 22035418, 2019--2022) funded by the ERA-NET
Cofund ForestValue. KM received funding from the Project No.
CZ.02.1.01/0.0/0.0/16_019/0000803 financed by OP RDE, grant no.
ITMS2014+313011W580 supported by the Integrated Infrastructure
Operational Programme funded by the ERDF. CH acknowledges financial
support from Waldfonds WASIM2100 project.
